---
title: "558_Project_1"
author: "Lee Bennett and Madelyn Garrison"
format: html
editor: visual
---

```{r}
#| include: false
#| echo: false

library(tidyverse)
library(tidycensus)
library(ggplot2)
```

# Retrieving and Processing Data from the PUMS Census API

First, we'll create a helper function that will process the raw data retrived from the API into a `tibble`. This function also adds a new class called `census` for this tibble, and we will use this class for summarizing the data later.

```{r}

cens_tibble <- function(rawdata) {
  pums_tib <- as_tibble(rawdata)
  
  class(pums_tib) <- c("census",class(pums_tib))
  
  return(pums_tib)
}
```


## Get PUMS data from Census API for a single year

```{r}

    trial<-function(nvar1='AGEP',nvar2=NULL,nvar3=NULL, nvar4=NULL, 
                    nvar5=NULL, nvar6=NULL,yr='2022', cvar1='SEX', 
                    cvar2=NULL, cvar3=NULL, cvar4=NULL, cvar5=NULL, 
                    cvar6=NULL,cvar7=NULL,cvar8=NULL,cvar9=NULL)
      {
              ends <- get_pums(variables = c(nvar1, nvar2, nvar3, nvar4, 
                                             nvar5, nvar6, cvar1, cvar2, 
                                             cvar3, cvar4, cvar5, cvar6, 
                                             cvar7, cvar8, cvar9),
                               survey = "acs1",
                               state = 'NC',
                               year = if (yr <= 2023 & yr >=2010){yr}
                              )
              return(ends)
      }

    x<-trial()
    
    tib_check <- pums_tbl(x)
```

Suppose a user wants to obtain multiple years of data for a given set of variables. The following function will call the query for a list of years specified as a vector:

```{r}

pums_years <- function(yr_list) {

}
```

# Exploring the PUMS Data

## Summary statistics

We previously created a class called `census` for tibbles generated from PUMS data queriers. We now create a custom summary function for tibbles of this class to produce means and standard deviations for selected numeric variables:

```{r}
summary.census <- function(cens_tibble,num_vars,cat_vars){
  
}
```


## Creating box plots

Finally, we create a function that allows us to generate a box plot for a specified numeric variable, grouped by the levels of a specified categorical variable:

```{r}
plot.census <- function(cens_tibble,num_var,cat_var) {
  ggplot(cens_tibble,aes(x=get(cat_var), y=get(num_var), weight=PWGTP)) + geom_boxplot()

}
```

